dyngen: a multi-modal simulator for spearheading new single-cell omics analyses

### 2020-09-15 manuscript/render.Rmd
move files around

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
    \caption{
      \textbf{Showcase of dyngen functionality.}
      \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
      \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
      \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
      \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth.
      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods \cite{saelens_comparisonsinglecelltrajectory_2019}.
      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
    }
    \label{fig:overview}
\end{figure}
dyngen was designed to include all of these functionalities and more by design. Its methodology allows tracking many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and knockdown experiments). The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases to simulate dynamic biological processes. The original design of dyngen was motivated by the plethora of methods available for trajectory inference, where dyngen allowed the first large-scale benchmarking of such methods [@saelens_comparisonsinglecelltrajectory_2019; @vandenberge_trajectorybaseddifferentialexpression_2020].

@@

# Availability
dyngen is available as an R package on GitHub at [github.com/dynverse/dyngen](https://github.com/dynverse/dyngen). The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).

@@

To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory. The geodesic distances between the pair of datasets are directly comparable to each other, and thus cells from the first trajectory should align relatively closely to cells from the second trajectory. We after aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.

@@

\begin{table}[H]
    \caption{
      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette. $^2$ As showcased by Van den Berge et al.  \cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
    } \label{tab:comparison}
    \small
    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
    \blackline
    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
    \blackline
    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
    \blackline
    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
    - & Reaction activity & \no & \no & \no & \no & \yes \\
    \blackline
    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
    \blackline
    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
    - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
    - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
    - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
    - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
    \blackline
    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
    \blackline
    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
    - & Time-series & \no & \no & \no & \no & \yes \\
    - & Snapshot & \no & \no & \no & \no & \yes \\
    \blackline
    \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
    \blackline
    - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
    - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
    \blackline
    %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
    %\blackline
    %- & Number of cells &  &  &  &  & \\ \grayline
    %- & Number of genes &  &  &  &  & \\ \grayline
    %- & Execution time & & & & & \\
    \blackline

\end{tabular}
\end{table}

>>>>

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
    \caption{
      \textbf{Showcase of dyngen functionality.}
      \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
      \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
      \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
      \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth.
      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods \cite{saelens_comparisonsinglecelltrajectory_2019}.
      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
    }
    \label{fig:overview}
\end{figure}

dyngen was designed to include all of these functionalities and more by design. Its methodology allows tracking many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions. dyngen can simulate a large variety of dynamic processes (e.g. cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g. batch effects and time-series, perturbation and knockdown experiments). The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases to simulate dynamic biological processes. The original design of dyngen was motivated by the plethora of methods available for trajectory inference, where dyngen allowed the first large-scale benchmarking of such methods [@saelens_comparisonsinglecelltrajectory_2019; @vandenberge_trajectorybaseddifferentialexpression_2020].

@@

# Availability
dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen). The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).

@@

To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory. The geodesic distances between the pair of datasets are directly comparable to each other, and thus cells from the first trajectory should align relatively closely to cells from the second trajectory. After aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.

@@

\begin{table}[H]
    \caption{
      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette. $^2$ As showcased by Van den Berge et al.  \cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
    } \label{tab:comparison}
    \small
    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
            \blackline
            & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
            \blackline
            \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
            \blackline
            - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
            - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
            - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
            - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
            - & Reaction activity & \no & \no & \no & \no & \yes \\
            \blackline
            \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
            \blackline
            - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
            - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
            - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
            - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
            - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
            - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
            - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
            - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
            \blackline
            \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
            \blackline
            - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
            - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
            - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
            - & Time-series & \no & \no & \no & \no & \yes \\
            - & Snapshot & \no & \no & \no & \no & \yes \\
            \blackline
            \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
            \blackline
            - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
            - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
            - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
            - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
            - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
            - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
            - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
            - & Trajectory alignment & \no & \no & \no & \no & \yes \\
            \blackline
            %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
            %\blackline
            %- & Number of cells &  &  &  &  & \\ \grayline
            %- & Number of genes &  &  &  &  & \\ \grayline
            %- & Execution time & & & & & \\
            \blackline
\end{tabular}
\end{table}
